---
title: "Getting Started"
author: "Peter Baumgartner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why This Package?

The Reporters Without Borders (RSF) Press Freedom Index measures press freedom
across countries annually. **pressfreedom.data** automates downloading,
cleaning, and standardizing this data for analysis in R.

The RSF index spans 24 years (2002-2026, with 2011 missing because RSF did not
publish that year) across three methodologically distinct periods:

- **2002-2012:** 16 columns, non-comparable scores
- **2013-2021:** 16 columns, comparable 0-100 scale
- **2022-2026:** 22-25 columns with new dimensions (Political, Economic,
  Legal, Social, Safety)

On top of the structural changes, countries have been renamed or
reclassified over time -- 14 pairs have been consolidated to a single,
current name, including:

- Turkey -> Turkiye (official name change, 2022)
- Ivory Coast -> Cote d'Ivoire (RSF standardization)
- Czech Republic -> Czechia (official name change)

Territorial variants are also resolved: Cyprus and Northern Cyprus are kept
as separate rows (they are distinct political entities), while other
territorial fragments (e.g., U.S. bases in Iraq) are removed. The audit
trail for these decisions -- which rows were consolidated and from what
original name -- is preserved in `data/processed/rwb_standardized.rds`
inside the package source, though it is not part of the exported
`rwb_standardized` dataset described below.

Every row in the exported dataset has a valid ISO 3166-1 alpha-3 country
code, so you can join it with other datasets without extra cleanup.

This package handles all of the above automatically, providing a single,
clean dataset ready for analysis.

## Installation

**From GitHub (recommended pre-CRAN):**

```r
pak::pak("petzi53/pressfreedom.data")

# or, using the classic alternative:
remotes::install_github("petzi53/pressfreedom.data")
```

**From source / local clone (development):**

```r
devtools::install()
```

**From CRAN** (once available):

```r
# install.packages("pressfreedom.data")
```

## Quick Start

### Load and Explore the Data

```{r load}
library(pressfreedom.data)

data(rwb_standardized)
dplyr::glimpse(rwb_standardized)
```

**Key columns:**

- `year_n` - Year (2002-2026; 2011 missing)
- `iso` - ISO 3166-1 alpha-3 country code
- `country_en` - Country name (English, ASCII-normalized)
- `score` - Press freedom score (0-100; lower = less free)
- `rank` - Ranking among countries that year (lower = more free)
- Dimension columns (available 2022+): `political_context`,
  `economic_context`, `legal_context`, `social_context`, `safety`
- Evolution columns: `rank_evolution`, `score_evolution` (change vs.
  previous year)

### Check Data Availability

```{r availability}
# Years available
unique(sort(rwb_standardized$year_n))

# Number of countries per year
rwb_standardized |>
  dplyr::group_by(year_n) |>
  dplyr::summarise(
    n_countries = dplyr::n_distinct(country_en),
    .groups = "drop"
  )
```

## Annual Updates

When RSF publishes new data (typically in spring), update the package with:

```r
# Auto-detects missing years, downloads, and processes
result <- update_rwb_data()

# View the result
print(result)
```

The function automatically:

1. Detects which years are missing
2. Downloads CSV files from RSF
3. Normalizes column names across periods
4. Combines into a unified structure
5. Standardizes country names and ISO codes
6. Validates the output
7. (Optionally) Commits changes to git

### Manual Control

For fine-grained control, you can run specific phases:

```r
# Skip download and cleaning, just recombine and standardize
result <- update_rwb_data(
  download = FALSE,
  clean = FALSE,
  standardize = TRUE,
  validate = TRUE
)

# Download and clean specific years
result <- update_rwb_data(
  years = c(2027),
  combine = TRUE,
  standardize = TRUE
)
```

See `?update_rwb_data` for all options and return values.

## FAQ & Troubleshooting

### Q: Why is 2011 missing?

**A:** RSF did not publish a Press Freedom Index for 2011. The package
preserves this gap rather than imputing values.

### Q: Can I use ISO codes to join with other datasets?

**A:** Yes. All rows have valid ISO 3166-1 alpha-3 codes in the `iso`
column. This is the standard for country joins.

### Q: Are dimensions available for all years?

**A:** No. The new dimensions (Political, Economic, Legal, Social, Safety)
are only available from 2022 onward. Earlier years have `NA` values.

### Q: What character encoding is used?

**A:** The data is normalized to ASCII, with special characters removed
(e.g., Cote d'Ivoire, Turkiye). This ensures compatibility across
platforms.

### Q: Are scores comparable across all 24 years?

**A:** No. RSF changed its scoring methodology in 2013, so scores from
2002-2012 are not directly comparable to 2013-2026. When analyzing trends
over time, restrict comparisons to the 2013-2026 period unless you are
specifically studying methodology effects.

### Q: How do I contribute corrections or suggestions?

**A:** Open an issue on GitHub: <https://github.com/petzi53/pressfreedom.data/issues>

### Q: Where can I learn more about the data source?

**A:** Visit the Reporters Without Borders Press Freedom Index:
<https://rsf.org/en/ranking>

### Q: Can I use this data in published research?

**A:** Yes. Please cite both the package and the original data source:

```
@software{baumgartner2026pressfreedom,
  author = {Baumgartner, Peter},
  title = {pressfreedom.data: Download and Process Reporters Without Borders
           Press Freedom Index Data},
  year = {2026},
  url = {https://github.com/petzi53/pressfreedom.data},
  note = {R package version 0.1.0}
}
```

And cite RSF:

```
@organization{rsf2026,
  title = {Press Freedom Index 2026},
  author = {Reporters Without Borders},
  year = {2026},
  url = {https://rsf.org/en/ranking}
}
```

## Summary

**pressfreedom.data** provides a clean, standardized press freedom dataset
spanning 24 years, with automatic handling of methodological changes,
country name consolidation, ISO codes, and annual update automation. For
chart-based examples, see the companion vignette
`vignette("visualizing-trends", package = "pressfreedom.data")`. For full
function documentation, visit the package website:
<https://www.peter-baumgartner.net/pressfreedom.data>.

Happy analyzing!
