---
title: "Getting Started with pressfreedom.data"
author: "Peter Baumgartner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pressfreedom.data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The Reporters Without Borders Press Freedom Index measures press freedom across countries annually. This vignette introduces **pressfreedom.data**, an R package that automates downloading, cleaning, and standardizing this data for analysis.

### Why This Package?

The RSF index spans 24 years (2002-2026) with evolving data structures:

- **2002-2012:** 16 columns, non-comparable scores
- **2013-2021:** 16 columns, comparable 0-100 scale  
- **2022-2026:** 22-25 columns with new dimensions (Political, Economic, Legal, Social, Safety)

Additionally, countries have been renamed (Turkey -> Turkey (now Turkiye), Ivory Coast -> Ivory Coast (now Cote d'Ivoire)), making historical analysis complex.

This package handles all these challenges automatically, providing a single, clean dataset ready for analysis.

## Installation

Install from source:

```r
devtools::install()
```

Or load the package directly:

```r
library(pressfreedom.data)
```

## Quick Start

### Load the Data

```{r load}
library(pressfreedom.data)

# Load the standardized dataset
data(rwb_standardized)

# View the first few rows
head(rwb_standardized, n = 10)
```

The dataset contains 4,192 rows representing 191 countries across 24 years:

```{r structure}
dim(rwb_standardized)
```

### Understanding the Data

Each row represents one country in one year:

```{r summary}
# View column descriptions
str(rwb_standardized)
```

**Key columns:**

- `year_n` - Year (2002-2026; 2011 missing)
- `iso` - ISO 3166-1 alpha-3 country code
- `country_en` - Country name (English, ASCII-normalized)
- `score` - Press freedom score (0-100; higher = less free)
- `rank` - Ranking among countries that year
- Dimension columns (available 2022+): `political_context`, `economic_context`, `legal_context`, `social_context`, `safety`
- Evolution columns: `rank_evolution`, `score_evolution` (change vs. previous year)

## Common Usage Patterns

### 1. Check Data Availability

```{r availability}
# Years available
unique(sort(rwb_standardized$year_n))

# Number of countries per year
rwb_standardized |>
  dplyr::group_by(year_n) |>
  dplyr::summarise(
    n_countries = dplyr::n_distinct(country_en),
    .groups = "drop"
  )
```

### 2. Track a Specific Country

```{r country_track}
# Track Germany's press freedom over time
germany <- rwb_standardized |>
  dplyr::filter(country_en == "Germany") |>
  dplyr::select(year_n, country_en, score, rank) |>
  dplyr::arrange(year_n)

head(germany)
```

### 3. Compare Countries in a Year

```{r year_compare}
# Top 10 most free countries in 2026 (lowest scores)
top_2026 <- rwb_standardized |>
  dplyr::filter(year_n == 2026) |>
  dplyr::slice_min(score, n = 10) |>
  dplyr::select(rank, country_en, score) |>
  dplyr::arrange(rank)

top_2026
```

### 4. Analyze Press Freedom Trends

```{r trends}
# Average press freedom score by decade
rwb_standardized |>
  dplyr::filter(!is.na(score)) |>
  dplyr::mutate(decade = floor(year_n / 10) * 10) |>
  dplyr::group_by(decade) |>
  dplyr::summarise(
    avg_score = mean(score),
    n_countries = dplyr::n_distinct(country_en),
    .groups = "drop"
  )
```

### 5. Work with Dimensions (2022+)

The dimensions (Political, Economic, Legal, Social, Safety) are only available from 2022 onward:

```{r dimensions}
# Check dimension availability
rwb_standardized |>
  dplyr::filter(year_n >= 2022) |>
  dplyr::select(
    year_n, country_en, 
    political_context, economic_context, legal_context, 
    social_context, safety
  ) |>
  head()
```

### 6. Regional Analysis

For earlier years, geographical zone data is available:

```{r zones}
# Zone distribution in 2010
rwb_standardized |>
  dplyr::filter(year_n == 2010, !is.na(zone)) |>
  dplyr::count(zone, sort = TRUE)
```

## Annual Updates

When RSF publishes new data (typically in spring), update the package with:

```r
# Auto-detects missing years, downloads, and processes
result <- update_rwb_data()

# View the result
print(result)
```

The function automatically:

1. Detects which years are missing
2. Downloads CSV files from RSF
3. Normalizes column names across periods
4. Combines into a unified structure
5. Standardizes country names and ISO codes
6. Validates the output
7. (Optionally) Commits changes to git

### Manual Control

For fine-grained control, you can run specific phases:

```r
# Skip download and cleaning, just recombine and standardize
result <- update_rwb_data(
  download = FALSE,
  clean = FALSE,
  standardize = TRUE,
  validate = TRUE
)

# Download and clean specific years
result <- update_rwb_data(
  years = c(2027),
  combine = TRUE,
  standardize = TRUE
)
```

See `?update_rwb_data` for all options and return values.

## Working with Analysis Packages

### Visualization with ggplot2

```r
library(ggplot2)

# Plot press freedom trends for selected countries
rwb_standardized |>
  dplyr::filter(country_en %in% c("Germany", "France", "United States", "China")) |>
  ggplot(aes(x = year_n, y = score, color = country_en)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Press Freedom Trends",
    x = "Year",
    y = "Score (0=most free, 100=least free)",
    color = "Country"
  ) +
  theme_minimal()
```

### Regression Analysis with broom

```r
library(broom)

# Does a country's rank improve or decline?
germany_model <- rwb_standardized |>
  dplyr::filter(country_en == "Germany", !is.na(score)) |>
  lm(score ~ year_n, data = _)

broom::tidy(germany_model)
```

### Time Series with tsibble

```r
library(tsibble)

# Convert to tsibble for time series analysis
ts_data <- rwb_standardized |>
  dplyr::filter(country_en == "Germany") |>
  dplyr::select(year_n, score) |>
  tsibble::as_tsibble(index = year_n)

ts_data
```

## Data Quality Notes

### Missing Years

Year 2011 is missing because RSF did not publish data that year. No imputation is performed.

### Country Consolidation

Fourteen country names have been consolidated to reflect official changes or RSF methodology updates:

- Turkey -> Turkey (now Turkiye)
- Ivory Coast -> Ivory Coast (now Cote d'Ivoire)
- Czech Republic -> Czechia
- And 11 others

Each consolidation is tracked in the `consolidated` column (TRUE if the row represents a consolidated entity).

### Territorial Variants

Cyprus and Northern Cyprus are kept separate because they are distinct political entities. Other territorial variants (e.g., U.S. bases in Iraq) have been removed.

### ISO Codes

All 4,192 rows have valid ISO 3166-1 alpha-3 country codes. You can use these for joining with other datasets.

## Accessing the Data

### Within R

Load the bundled dataset:

```r
data(rwb_standardized)
```

### Direct File Access

For external use, the standardized dataset is also available as an R data file:

```r
# Path to the data file
system.file("data", "rwb_standardized.rda", package = "pressfreedom.data")
```

## FAQ & Troubleshooting

### Q: Why is 2011 missing?

**A:** RSF did not publish a Press Freedom Index for 2011. The package preserves this gap rather than imputing values.

### Q: Can I use ISO codes to join with other datasets?

**A:** Yes. All rows have valid ISO 3166-1 alpha-3 codes in the `iso` column. This is the standard for country joins.

### Q: Are dimensions available for all years?

**A:** No. The new dimensions (Political, Economic, Legal, Social, Safety) are only available from 2022 onward. Earlier years have NA values.

### Q: What character encoding is used?

**A:** The data is normalized to ASCII, with special characters removed (e.g., Ivory Coast (now Cote d'Ivoire), Turkey (now Turkiye)). This ensures compatibility across platforms.

### Q: How do I contribute corrections or suggestions?

**A:** Open an issue on GitHub: https://github.com/petzi53/pressfreedom.data/issues

### Q: Where can I learn more about the data source?

**A:** Visit the Reporters Without Borders Press Freedom Index: https://rsf.org/en/ranking

### Q: Can I use this data in published research?

**A:** Yes. Please cite both the package and the original data source:

```
@software{baumgartner2026pressfreedom,
  author = {Baumgartner, Peter},
  title = {pressfreedom.data: Download and Process Reporters Without Borders 
           Press Freedom Index Data},
  year = {2026},
  url = {https://github.com/petzi53/pressfreedom.data},
  note = {R package version 0.1.0}
}
```

And cite RSF:

```
@organization{rsf2026,
  title = {Press Freedom Index 2026},
  author = {Reporters Without Borders},
  year = {2026},
  url = {https://rsf.org/en/ranking}
}
```

## Summary

**pressfreedom.data** provides:

[OK] Clean, standardized press freedom data for 24 years (2002-2026)  
[OK] Automatic handling of methodological changes  
[OK] Country name consolidation and ISO codes  
[OK] Annual update automation  
[OK] Full data pipeline transparency  

For more details, see the function documentation:

```r
?update_rwb_data
?rwb_standardized
?download_rwb_data
?standardize_rwb_countries
```

Happy analyzing!
