---
title: "Visualizing Press Freedom Trends"
author: "Peter Baumgartner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Press Freedom Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk-options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Purpose

This vignette is a gallery, not an analysis. Its only goal is to show, through a handful of charts, the kinds of questions {pressfreedom.data} makes possible to ask. For instance: 

- tracking a single country over time, 
- comparing countries against each other, 
- breaking a score down into its component dimensions
- following global and regional trends.

None of the charts below are meant to support a substantive conclusion about press freedom; even if some conclusions are not far to seek. Read these charts as demonstrations of the data's shape and coverage, aimed at researchers deciding whether this dataset is useful for their own work. For a conceptual introduction to the dataset (its columns, time periods, and known data-quality caveats), see vignette("getting-started").

## Setup

The charts in this vignette use {ggplot2} for standard line/point charts,
{ggbump} for rank-crossing ("bump") charts, and {tidyr} for reshaping
wide dimension columns into a long format for plotting. {dplyr} is used
throughout for filtering, grouping, and summarizing.

```{r setup}
library(pressfreedom.data)
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggbump)
library(patchwork)

data(rwb_standardized)
```

**A reminder before plotting:** 

- `score` is only comparable from 2013 onward -- RSF
changed its scoring methodology that year. 
- `rank` is always comparable
across the full 2002-2026 span. It is just a country's position among that
year's countries.
- The five dimension columns (`political_context`, `economic_context`, `legal_context`, `social_context`, and `safety`) are available from 2022 onward.
- The year 2011 is completely missing. RSF.org did not publish data for that year.

## Track a Specific Country

`rank` and `score` tell complementary stories for a single country. `rank`
is a country's position among that year's countries, so it is comparable
across the whole 2002-2026 span. `score` uses RSF's underlying point scale,
which is only comparable from 2013 onward. The two charts below show both
for the United States, side by side, each starting at the year from which
its values are meaningful.

```{r us-setup}
us_rank <- rwb_standardized |>
  filter(country_en == "United States", !is.na(rank)) |>
  arrange(year_n)

us_score <- rwb_standardized |>
  filter(country_en == "United States", year_n >= 2013, !is.na(score)) |>
  arrange(year_n)
```

```{r us-charts, fig.alt = "Two line charts side by side: the United States' press freedom rank from 2002 to 2026 on the left, and its score from 2013 to 2026 on the right."}
p_us_rank <- ggplot(us_rank, aes(x = year_n, y = rank)) +
  geom_line() +
  geom_point() +
  scale_y_reverse() +
  labs(x = "Year", y = "Rank (1 = most free)", title = "Rank, 2002-2026")

p_us_score <- ggplot(us_score, aes(x = year_n, y = score)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Score (100 = most free)",
    title = "Score, 2013-2026"
  )

p_us_rank + p_us_score
```

## Compare Countries Over Time

The six countries below -- the United States, China, Brazil,
Nigeria, Japan, and Germany -- span a mix of regions and press-freedom
trajectories.

```{r compare-setup}
compare_countries <- c(
  "United States", "China", "Brazil", "Nigeria", "Japan", "Germany"
)

compare_score <- rwb_standardized |>
  filter(country_en %in% compare_countries, year_n >= 2013, !is.na(score)) |>
  arrange(country_en, year_n)

compare_rank <- rwb_standardized |>
  filter(country_en %in% compare_countries, !is.na(rank)) |>
  arrange(country_en, year_n)
```

```{r compare-score-chart, fig.alt = "Line chart comparing press freedom scores for the United States, China, Brazil, Nigeria, Japan, and Germany from 2013 to 2026."}
ggplot(compare_score, aes(x = year_n, y = score, color = country_en)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Score (100 = most free)",
    color = "Country",
    title = "Press Freedom Score, 2013-2026"
  )
```

For rank, a bump chart (`ggbump::geom_bump()`) shows how the six countries'
relative positions cross over the full 2002-2026 span, since `rank` does not
share `score`'s 2013 comparability limit:

```{r compare-rank-chart, fig.alt = "Bump chart comparing press freedom rank for the United States, China, Brazil, Nigeria, Japan, and Germany from 2002 to 2026, with a reversed y-axis.", fig.height = 6}
ggplot(compare_rank, aes(x = year_n, y = rank, color = country_en)) +
  ggbump::geom_bump(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_reverse() +
  labs(
    x = "Year",
    y = "Rank (1 = most free)",
    color = "Country",
    title = "Press Freedom Rank, 2002-2026"
  )
```

## Work with Dimensions (2022+)

From 2022 onward, RSF also reports five sub-dimensions (Political, Economic,
Legal, Social, Safety) alongside the overall `score`. Reshaping with
`tidyr::pivot_longer()` makes it easy to plot them together for one country.

```{r dimensions_setup}
us_dims <- rwb_standardized |>
  filter(country_en == "United States", year_n >= 2022) |>
  select(
    year_n, score, political_context, economic_context,
    legal_context, social_context, safety
  ) |>
  tidyr::pivot_longer(
    cols = -"year_n",
    names_to = "dimension",
    values_to = "value"
  )
```

```{r dimensions_chart, fig.alt = "Line chart for the United States' overall score and five sub-dimensions from 2022 to 2026."}
ggplot(us_dims, aes(x = year_n, y = value, color = dimension)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Score (higher is better)",
    color = "Dimension",
    title = "United States: Overall Score and Sub-Dimensions, 2022-2026"
  )
```

## Regional Trends



Averaging `score` by `zone` and `year_n` (restricted to 2013+, since `score`
is not comparable before then) shows how regions have diverged. The black
line adds the global mean across all countries for reference.

```{r zones_setup}
zone_means <- rwb_standardized |>
  filter(year_n >= 2013, !is.na(zone), !is.na(score)) |>
  group_by(zone, year_n) |>
  summarise(mean_score = mean(score), .groups = "drop")

global_mean <- rwb_standardized |>
  filter(year_n >= 2013, !is.na(score)) |>
  group_by(year_n) |>
  summarise(mean_score = mean(score), .groups = "drop")
```

```{r zones_chart, fig.alt = "Line chart of mean press freedom score by geographic zone from 2013 to 2026, with a black line showing the global mean across all countries."}
zone_colors <- c(
  setNames(scales::hue_pal()(dplyr::n_distinct(zone_means$zone)), sort(unique(zone_means$zone))),
  "Global mean" = "black"
)

ggplot(mapping = aes(x = year_n, y = mean_score, color = zone)) +
  geom_line(data = zone_means) +
  geom_point(data = zone_means) +
  geom_line(
    data = global_mean,
    aes(color = "Global mean"),
    linewidth = 1
  ) +
  geom_point(data = global_mean, aes(color = "Global mean")) +
  scale_color_manual(values = zone_colors) +
  labs(
    x = "Year",
    y = "Mean score",
    color = "Zone",
    title = "Mean Press Freedom Score by Region, 2013-2026"
  )
```

## The Six RSF Regions

`zone` groups countries into six regions, following RSF's own regional
breakdown. These are RSF's groupings, not a universal standard. Note, for instance, that North Africa sits in `Middle East & North Africa` rather than `Africa`. RSF's regions
follow political/cultural groupings, not strict continental geography.
But nothing stops a researcher from building alternative regions (e.g. by continent, by income group, by EU membership) directly from `country_en` or `iso`.

- **Africa -- Sub-Saharan Africa:** Angola, Benin, Botswana, Burkina Faso,
  Burundi, Cabo Verde, Cameroon, Central African Republic, Chad, Comoros,
  Congo-Brazzaville, Cote d'Ivoire, DR Congo, Djibouti, Equatorial Guinea,
  Eritrea, Eswatini, Ethiopia, Gabon, Gambia, Ghana, Guinea, Guinea-Bissau,
  Kenya, Lesotho, Liberia, Madagascar, Malawi, Mali, Mauritania, Mauritius,
  Mozambique, Namibia, Niger, Nigeria, Rwanda, Senegal, Seychelles, Sierra
  Leone, Somalia, South Africa, South Sudan, Sudan, Tanzania, Togo, Uganda,
  Zambia, Zimbabwe.
- **Americas -- North, Central, and South America, and the Caribbean:**
  Argentina, Belize, Bolivia, Brazil, Canada, Chile, Colombia, Costa Rica,
  Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Guyana, Haiti,
  Honduras, Jamaica, Mexico, Nicaragua, OECS, Panama, Paraguay, Peru,
  Suriname, Trinidad and Tobago, United States, Uruguay, Venezuela.
- **Asia-Pacific -- Asia and the Pacific:** Afghanistan, Australia,
  Bangladesh, Bhutan, Brunei, Cambodia, China, East Timor, Fiji, Hong Kong,
  India, Indonesia, Japan, Laos, Malaysia, Maldives, Mongolia, Myanmar,
  Nepal, New Zealand, North Korea, Pakistan, Papua New Guinea, Philippines,
  Samoa, Singapore, South Korea, Sri Lanka, Taiwan, Thailand, Tonga,
  Vietnam.
- **Eastern Europe & Central Asia -- Commonwealth of Independent States:**
  Armenia, Azerbaijan, Belarus,
  Georgia, Kazakhstan, Kyrgyzstan, Moldova, Russia, Tajikistan, Turkiye,
  Turkmenistan, Ukraine, Uzbekistan.
- **Middle East & North Africa:** Algeria,
  Bahrain, Egypt, Iran, Iraq, Israel, Jordan, Kuwait, Lebanon, Libya, Morocco
  / Western Sahara, Oman, Palestine, Qatar, Saudi Arabia, Syria, Tunisia,
  United Arab Emirates, Yemen.
- **EU & Balkans -- European Union member states, EFTA countries, the UK,
  and the Balkans:** Albania, Andorra, Austria, Belgium, Bosnia-Herzegovina,
  Bulgaria, Croatia, Cyprus, Czechia, Denmark, Estonia, Finland, France,
  Germany, Greece, Hungary, Iceland, Ireland, Italy, Kosovo, Latvia,
  Liechtenstein, Lithuania, Luxembourg, Malta, Montenegro, Netherlands,
  North Macedonia, Northern Cyprus, Norway, Poland, Portugal, Romania,
  Serbia, Slovakia, Slovenia, Spain, Sweden, Switzerland, United Kingdom.

