---
title: "OLD REFERENCE -- Visualizing Press Freedom Trends (not built, not shipped)"
author: "Peter Baumgartner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

> **This file is a reference copy only.** It is being replaced by a
> rewritten `vignettes/visualizing-trends.Rmd`. The `%\Vignette...` engine
> tags have been removed so R CMD check does not try to build it as a
> vignette. Once the new vignette is finished, move this file out of
> `vignettes/` (e.g. into `.posit/assistant/docs/`) or delete it.

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette walks through common charts built on `rwb_standardized`. Each
example uses **ggplot2**, with **ggbump** for rank-crossing charts and
**tidyr** for reshaping. See `vignette("getting-started")` for an
introduction to the dataset and its columns.

```{r load}
library(pressfreedom.data)
library(dplyr)
library(ggplot2)

data(rwb_standardized)
```

**A reminder before plotting:** `score` (and, from 2022 on, the dimension
columns) is only comparable from 2013 onward -- RSF changed its scoring
methodology that year. Charts below that compare scores across years filter
to `year_n >= 2013`.

## Track a Specific Country

`rank` and `score` tell complementary stories: `rank` is always comparable
across the full 2002-2026 span (it is just a country's position among that
year's countries), while `score` is only comparable within the 2013+ period.

```{r germany_setup}
germany <- rwb_standardized |>
  filter(country_en == "Germany") |>
  arrange(year_n)
```

```{r germany_score, fig.alt = "Line chart of Germany's press freedom score from 2002 to 2026, showing a sharp jump around 2013."}
ggplot(germany, aes(x = year_n, y = score)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Score (0 = most free, 100 = least free)",
    title = "Germany: Press Freedom Score Over Time"
  )
```

The visible jump between 2012 and 2013 is the methodology change, not a real
improvement in press freedom -- another reminder not to read this chart's
pre-2013 and post-2013 halves as one continuous trend.

`rank` does not have this problem, since it is always relative to the other
countries scored that year:

```{r germany_rank, fig.alt = "Line chart of Germany's press freedom rank from 2002 to 2026 with a reversed y-axis so rank 1 is at the top."}
ggplot(germany, aes(x = year_n, y = rank)) +
  geom_line() +
  geom_point() +
  scale_y_reverse() +
  labs(
    x = "Year",
    y = "Rank (1 = most free)",
    title = "Germany: Press Freedom Rank Over Time"
  )
```

## Compare Countries Over Time

To compare multiple countries, filter to the comparable-score period
(2013-2026) and map `color` to `country_en`.

```{r compare_setup}
countries <- c("Germany", "France", "United States", "China")

compare_df <- rwb_standardized |>
  filter(country_en %in% countries, year_n >= 2013) |>
  arrange(country_en, year_n)
```

```{r compare_score, fig.alt = "Line chart comparing press freedom scores for Germany, France, the United States, and China from 2013 to 2026."}
ggplot(compare_df, aes(x = year_n, y = score, color = country_en)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Score (0 = most free, 100 = least free)",
    color = "Country",
    title = "Press Freedom Score, 2013-2026"
  )
```

For rank, a bump chart (`ggbump::geom_bump()`) is a better fit than a plain
line chart, since it is purpose-built for showing how several entities'
ranks cross over time:

```{r compare_rank, fig.alt = "Bump chart comparing press freedom rank for Germany, France, the United States, and China from 2013 to 2026, with a reversed y-axis."}
ggplot(compare_df, aes(x = year_n, y = rank, color = country_en)) +
  ggbump::geom_bump(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_reverse() +
  labs(
    x = "Year",
    y = "Rank (1 = most free)",
    color = "Country",
    title = "Press Freedom Rank, 2013-2026"
  )
```

## Analyze Global Trends

Averaging `score` across all countries by year -- restricted to 2013-2026 for
the same comparability reason -- gives a sense of the global trend.

```{r global_trend_table}
yearly_mean <- rwb_standardized |>
  filter(year_n >= 2013, !is.na(score)) |>
  group_by(year_n) |>
  summarise(
    mean_score = mean(score),
    n_countries = n_distinct(country_en),
    .groups = "drop"
  )

yearly_mean
```

```{r global_trend_chart, fig.alt = "Line chart of the mean press freedom score across all countries from 2013 to 2026, showing a decline especially from 2022 onward."}
ggplot(yearly_mean, aes(x = year_n, y = mean_score)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Mean score across all countries",
    title = "Global Mean Press Freedom Score, 2013-2026"
  )
```

The mean score declines gradually through the 2010s and then more steeply
from 2022, coinciding with RSF's introduction of the five sub-dimensions
that year -- worth keeping in mind as a possible methodology effect rather
than treating the whole 2013-2026 span as a single comparable trend.

## Work with Dimensions (2022+)

From 2022 onward, RSF also reports five sub-dimensions (Political, Economic,
Legal, Social, Safety) alongside the overall `score`. Reshaping with
`tidyr::pivot_longer()` makes it easy to plot them together for one country.

```{r dimensions_setup}
germany_dims <- rwb_standardized |>
  filter(country_en == "Germany", year_n >= 2022) |>
  select(
    year_n, score, political_context, economic_context,
    legal_context, social_context, safety
  ) |>
  tidyr::pivot_longer(
    cols = -"year_n",
    names_to = "dimension",
    values_to = "value"
  )
```

```{r dimensions_chart, fig.alt = "Line chart of Germany's overall score and five sub-dimensions from 2022 to 2026."}
ggplot(germany_dims, aes(x = year_n, y = value, color = dimension)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Score",
    color = "Dimension",
    title = "Germany: Overall Score and Sub-Dimensions, 2022-2026"
  )
```

## Regional Analysis

`zone` groups countries into six RSF-defined regions. Averaging `score` by
`zone` and `year_n` (again restricted to 2013+) shows how regions have
diverged.

```{r zones_setup}
zone_means <- rwb_standardized |>
  filter(year_n >= 2013, !is.na(zone), !is.na(score)) |>
  group_by(zone, year_n) |>
  summarise(mean_score = mean(score), .groups = "drop")
```

```{r zones_chart, fig.alt = "Line chart of mean press freedom score by geographic zone from 2013 to 2026."}
ggplot(zone_means, aes(x = year_n, y = mean_score, color = zone)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Mean score",
    color = "Zone",
    title = "Mean Press Freedom Score by Region, 2013-2026"
  )
```

MENA and EEAC show the steepest declines from 2022 onward, while UE Balkans
remains the highest-scoring region throughout -- though as above, some of the
post-2022 shift across all zones may reflect RSF's 2022 methodology change
rather than a real change in conditions.
